---
title: "The color of roads"
description: |
  What colors are most represented in the road names of Dane County?
author:
  - name: Harald Kliems 
    orcid_id: 0000-0003-2542-0047
    url: https://haraldkliems.netlify.app/
date: "2022-11-19"
creative_commons: CC BY-SA
categories:
  - Madison (WI)
  - Dane County (WI)
  - map
output: 
  distill::distill_article:
    code_folding: TRUE
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

Sometimes you start a project and it ends up quite different from you had envisioned. I was going to do a quick and simple map for the [#30DayMapChallenge](https://30daymapchallenge.com/) theme "Blue." Pull road data for Dane County, filter for names that contain "blue," plot to a map with roads and road names. Sounds simple, right? Of course it is not. 


Getting all the data in is straightforward with `tigris` and `tmaptools`.

```{r}
library(tigris)
options(tigris_use_cache = TRUE)
library(tmap)
library(tmaptools)
library(sf)
library(tidyverse)
library(extrafont)
loadfonts(device = "all")

wi_roads <- roads("WI", "Dane")

# county shapefile to calculate bounding box for basemap download
dane <- tigris::counties("WI", cb = T) %>% 
  filter(NAME == "Dane")

#get basemap from OSM
dane_osm <- read_osm(bb(dane), zoom = 10, type = "stamen-toner", )

blue_dane <- wi_roads |> 
  mutate(blue = ifelse(str_detect(FULLNAME, "[Bb]lue"), "blue", "not blue")) |> 
  filter(blue == "blue") #keep only blue roads

blue_dane
```

But a map shows that more work is needed:

```{r layout="l-body-outset"}
tm_shape(dane_osm) +
  tm_rgb() +
  tm_shape(blue_dane) +
  tm_lines("blue", lwd = 2) +
  tm_text("FULLNAME", col = "blue")
```

Not good: The labels are too crowded, and you can see that a single way with the same name often is split into multiple segments, e.g. Blue Mounds Trail on the western edge of the map. I spent hours trying to fix these things through one of the various geometric or other tools from the `sf` package. But in the end, nothing worked. (If you have suggestions on how to combine adjoining ways, I'm all ears!) 

Instead, I pivoted to making a map of _all_ the road name colors in Dane County. The colors are extracted with simple regular expressions. This works well for all colors except "red" -- RiveREDege Ct, LaREDo Ct, or EldRED St are just some of the false positives. For one county these can be cleaned manually, but for a larger dataset this would be a problem.

```{r}
# deal with false-positive reds
not_red <- c("Riveredge Rd", 
             "Laredo Ct",
             "Eldred St", 
             "Fredericksburg Ct", 
             "Frederick Ct", 
             "Arboredge Way", 
             "Frederick Cir",
             "Redan Dr",
             "Meredith Way",
             "Claredon Dr",
             "Frederick St",
             "Fredericksburg Ln",
             "Meredithe Ave",
             "Arboredge Way",
             "Fredenberg Rd",
             "Mildred Ct",
             "Hubred Ln",
             "Covered Bridge Trl",
             "Saddle Bred Ln",
             "Saddlebred Ln")

blue_dane <- wi_roads |> 
  mutate(blue = case_when(str_detect(FULLNAME, "[Bb]lue") ~ "blue",
                          str_detect(FULLNAME, "[Bb]lack") ~ "black",
                          str_detect(FULLNAME, "[Rr]ed") ~ "red",
                          str_detect(FULLNAME, "[Yy]ellow") ~ "yellow",
                          str_detect(FULLNAME, "[Bb]rown") ~ "brown",
                          str_detect(FULLNAME, "[Gg]r[ae]y") ~ "grey",
                          str_detect(FULLNAME, "[Gg]reen") ~ "green",
                          T ~ NA_character_)) |> 
  filter(!is.na(blue) & !blue %in% not_red)
```

Which color is most common?

```{r}
blue_dane |> 
  st_drop_geometry() |>
  group_by(blue) |> 
  tally(sort = T) |> 
  knitr::kable()
```

Green! And here are the two maps, one dynamic and one static:

```{r}
tmap_mode("view")
tm_shape(blue_dane) +
  tm_lines("blue",lwd = 3, alpha = .8,  id = "FULLNAME") +
  tm_layout(title = "The Color of Dane County Road Names")
   
```


```{r layout="l-body-outset"}
tmap_mode("plot")
tm_shape(dane_osm) +
  tm_rgb(alpha = .7) +
  tm_shape(blue_dane) +
  tm_lines("blue",lwd = 3, alpha = .8) +
  tm_layout(title = "The Color of Dane County Road Names", fontfamily = "Roboto Condensed", title.bg.color = "lightgrey", title.bg.alpha = .8)
```


