---
title: "#30DayMapChallenge: Restaurant heat map of Madison"
description: |
  Challenge 15: Food/drink
author:
  - name: Harald Kliems 
    orcid_id: 0000-0003-2542-0047
    url: https://haraldkliems.netlify.app/
date: "2022-11-16"
creative_commons: CC BY-SA
categories:
  - Madison (WI)
  - map
output: 
  distill::distill_article:
    code_folding: TRUE
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

```{r}
library(tidyverse)
library(sf)
library(tmap)
library(mapboxapi)
library(leaflet)
library(leaflet.extras)


```

Another quick [#30DayMapChallenge](https://30daymapchallenge.com/) post. The prompt for day 15: Food/drink. Public Health Madison & Dane County have [a dataset](https://data-cityofmadison.opendata.arcgis.com/datasets/licensed-establishment/explore) on health inspections which includes all licensed restaurants in the county. The data are not geocoded, and so I use the `mapboxapi` package to geocode  the locations:


```{r}
establishments <- read_csv("data/Licensed_Establishment.csv")

# geocode and save geocoded data
# est_sf <- establishments %>% 
#   rowwise() %>% 
#   mutate(geometry = mb_geocode(AddressFull, output = "sf"),
#          geometry = geometry$geometry)
# 
# est_sf <- st_sf(est_sf)
# 
# write_rds(est_sf, file = "data/licensed_establishment_geocoded.RDS")
est_sf <- readRDS("data/licensed_establishment_geocoded.RDS")
```

The dataset includes "all operational licensed establishments receiving routine inspections," that is, not just restaurants. A quick `tmap` (we have to filter out a food cart that's located in Milwaukee):

```{r layout="l-body-outset"}
tmap_mode("plot")
est_sf %>% 
  filter(AddrCity != "MILWAUKEE") %>% 
  tm_shape() +
  tm_dots("EstablishmentType") +
  tm_layout(legend.outside = T)
```


Now we filter to only establishments that are of the type "Primarily Restaurant" and create a heat map. Heat maps look cool, but I admittedly don't fully understand how `leaflet.extras` generates them under the hood -- certainly the map looks very different depending on the zoom level. 

``` {r layout="l-body-outset"}
leaflet(est_sf %>% filter(EstablishmentType == "Primarily Restaurant")) %>% 
  addTiles() %>% 
  addHeatmap(radius = 6)
```
