---
title: "Some furter zoning analyses for Madison"
description: |
  A follow-up post on zoning and density in Madison and how it compares to other cities in the US.
author:
  - name: Harald Kliems
    url: https://haraldkliems.netlify.app/
date: 06-25-2019
tags:
  - R
  - zoning
  - landuse
  - Madison
output:
  distill::distill_article:
    self_contained: false
---
```{r}
library(tidyverse)
library(kableExtra)
```

I got some good feedback after posting my article [on Madison's zoning restrictions](https://haraldkliems.netlify.app/posts/how-much-of-madison-allows-only-detached-single-family-housing/) on Facebook and Reddit, prompting some additional analyses.

# Why do you look only at residentially zoned land?

Several people pointed out that comparing the zoning districts that only allow detached single-family homes (SFR) to all *residential* zoning districts may be misleading. In many cities the densest parts are not in residential zoning districts but in downtown areas zoned for mixed use. I think that's a fair point, but it was the original NYT article that chose only residential areas as the comparator, and I wanted to make Madison comparable to the other cities in the NYT analysis. 

Nonetheless, I quickly ran the analysis comparing SFR to any district except "special districts." These includes parks, conservancies, agricultural land, airports, or the UW-Campus---in other words districts that are highly unlikely to ever have residential use. 

```{r}

data <- read_csv(
  "https://opendata.arcgis.com/datasets/4bfd0deffa8f4d5f8eefe868ab91493c_9.csv",
  col_types = "iccccccdd")

sfr_zones <- c("SR-C1", 
         "SR-C2",
         "TR-C1", 
         "TR-C2", 
         "TR-C3", 
         "TR-R")

other <- c("A",
           "UA",
           "CN",
           "PR",
           "AP",
           "CI",
           "PD",
           "PHMP")
```
```{r echo=T}
#detached single-family only
res_sfr <- data %>%
  filter(ZONING_CODE %in% sfr_zones) %>%
  summarize(sum(ShapeSTArea))
# only special district area
other_area <- data %>%
  filter(ZONING_CODE %in% other) %>%
  summarise(sum(ShapeSTArea))
total <- data %>% summarize(sum(ShapeSTArea))
res_sfr/(total-other_area)
```

The result? 49% of all land that reasonably could have residential units on it allows only single-family attached housing. Since there is no point of comparison to other cities, it's hard to interpret that number.

# Is there a correlation between the percentage of SFR zoning and overall density?

It's a good question: Does the metric the NYT used actually have an association with overall density of a city? (Note that even if there is a correlation, this absolutely doesn't mean there is a causal connection! There are many, many factors that impact a city's overall density.)

There are few data points to work with. I retrieved density information for all cities in the NYT article and Madison from Wikipedia:

```{r}
dens_zon <- read_csv("data/density_zoning.csv")

dens_zon %>% 
  pivot_wider(names_from = Measure, values_from = Value) %>% 
  mutate(sfr = sfr *100) %>% 
  arrange(desc(density)) %>% 
  kable(digits = 0, col.names = c("City", "Density (pop./sq.mi.)", "Single-family (%)"))
```

Let's see what the overall correlation is:

```{r}
 dens_zon %>% 
   spread(Measure, Value) %>%
   summarize(cor(density, sfr))
```

Hm, -0.89. That's a very, very high correlation. But it's always good to look at your data points in a scatterplot to see what's actually going on:
 
```{r preview=T}
library(ggrepel)
library(gghighlight)
 dens_zon %>% 
   spread(Measure, Value) %>%
   ggplot(aes(sfr, density, label = City)) +
   geom_point() +
   geom_text_repel() +
   xlab("Proportion detached single family zoning to all residential zoning") +
   ylab("Overall density (population/sq mi)")
```


Okay, New York is clearly way out there. Let's exclude NYC and do the correlation again:

```{r echo=T}
 dens_zon %>% 
   spread(Measure, Value) %>%
   filter(City != "New York") %>%
   summarize(cor(density, sfr))
```


That gets us a correlation of -0.69. Much lower, but still pretty high. Is DC also an outlier?



```{r echo=T}
 dens_zon %>% 
   spread(Measure, Value) %>%
   filter(City != "New York" & City != "Washington") %>%
   summarize(cor(density, sfr))
```

This lower the correlation to -0.37. That is, there is a small negative association between how dense a city is and how much of its residential land is zoned exclusively for detached single-family homes. But more data is needed to confirm this. 