---
title: "Understanding Adjusted Data in the National Transit Database"
description: |
  A short description of the post.
author:
  - name: Harald Kliems
    url: https://haraldkliems.netlify.app/
date: 2023-03-26
output:
  distill::distill_article:
    self_contained: false
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

I regularly post about transit data, here and on social media -- usually about ridership for Madison Metro. Those posts are always based on data from [National Transit Database (NTD)](https://www.transit.dot.gov/ntd/ntd-data). One thing I kept encountering were discrepancies in the ridership data: I saw one number in the NTD spreadsheets and another number mentioned in, say, a local press conference. 

There are two types of monthly data available from NTD: ["adjusted"](https://www.transit.dot.gov/ntd/data-product/monthly-module-adjusted-data-release) and ["raw"]() data. 


I spent a good amount of time trying to find out what exactly NTD mean by adjusted data. On their download page they say this (bolding mine):

> "Contains monthly-updated service information reported by urban Full Reporters, **released after adjustments.**

> This file will be updated periodically by FTA to include data for subsequent months, and to **incorporate revisions to prior months** in the calendar year made by the transit properties. In some cases, this may include **revisions to data from previous years.**

> Any data value that appears in this file that does not appear in the Raw Database Time Series is based on a **growth estimate developed by FTA.** The strength of each estimate improves when the population of agencies reporting in the prior period is larger. "

A [PowerPoint presentation on using NTD data](https://www.transit.dot.gov/ntd/ntd-data-users-guide), states that "Adjusted Monthly Module Data Release Is Adjusted to Support Time-Series Study." And cautions, "Data from the most-recent Month may be incomplete due to late reporters. Data is subject to change for up to 18 months (or more in rare cases)." 

None of this is very helpful in understanding _how_ the adjusted data are being adjusted. For an agency like Metro, which as a full reporter is required to report data monthly, it would seem to make sense to just use the raw data. Let's compare raw and adjusted:

```{r}
library(tidyverse)
library(ntdr)
metro_raw <- get_ntd(agency = "City of Madison", data_type = "raw", modes = "MB", cache = TRUE)
metro_adjusted <- get_ntd(agency = "City of Madison", data_type = "adjusted", modes = "MB", cache = TRUE)

metro_raw <- metro_raw |> 
  mutate(data_type = "raw")
metro_adjusted <- metro_adjusted |> 
  mutate(data_type = "adjusted")

metro_all <- rbind(metro_raw, metro_adjusted)

ggplot(metro_all, aes(month, value, color = data_type)) +
  geom_line(alpha = .5)
```

The raw data is not available for the most recent year. Let's look at the ReadMe tab of the raw data release:

The Monthly module includes a limited set of key indicators reported by transit properties. Data is reported on a monthly basis, by mode and type of service, for a fiscal year." And further down: 
"Unlinked Passenger Trips – For the most recent closed-out annual report year."

So the raw data only seem to be made available after a full year has been closed out. This is really confusing.

Back to the adjusted data. Because I'm really bad at cleaning out my "Downloads" folder, I have several monthly adjusted data files released in different months in 2022. This gives us the ability to compared the files and see how large the adjustments being made actually are.

```{r}

library(readxl)
bus_dec_22 <- read_excel("c:/Users/user1/Downloads/December 2022 Adjusted Database(1).xlsx", sheet = 3)

bus_oct_22 <- read_excel("C:/Users/user1/Downloads/October 2022 Adjusted Database(1).xlsx", sheet = 3)

bus_mar_22 <- read_excel("c:/Users/user1/Downloads/March 2022 Adjusted Database.xlsx", sheet = 3)

bus_long_dec_22 <- bus_dec_22 %>% 
  filter(Agency == "City of Madison" & Modes == "MB") |> 
  pivot_longer(cols = 10:ncol(bus_dec_22), names_to = "month", values_to = "value") %>% 
  mutate(date = my(month),
         month = month(date),
         report_month = "dec_22")


bus_long_mar_22 <- bus_mar_22 %>% 
  filter(Agency == "City of Madison" & Modes == "MB") |> 
  pivot_longer(cols = 10:ncol(bus_mar_22), names_to = "month", values_to = "value") %>% 
  mutate(date = my(month),
         month = month(date),
         report_month = "mar_22")

bus_long_oct_22 <- bus_oct_22 %>% 
  filter(Agency == "City of Madison" & Modes == "MB") |> 
  pivot_longer(cols = 10:ncol(bus_oct_22), names_to = "month", values_to = "value") %>% 
  mutate(date = my(month),
         month = month(date),
         report_month = "oct_22")

metro_comp <- rbind(bus_long_dec_22, bus_long_oct_22, bus_long_mar_22)

metro_comp|> 
  ggplot(aes(date, value, color = report_month)) +
  geom_line() +
  title(main = "Comparing NTD monthly adjusted UPT data released in different months")
```

The three lines overlap nicely for any data prior to about 2020. Let's zoom in on the later data:

```{r}
metro_comp |> 
  filter(date > ymd("2020-01-01")) |> 
  ggplot(aes(date, value, color = report_month)) +
  geom_line()
  
```

The most obvious discrepancy is between the data released in March 2022 and the October and December releases: The March data starts diverging in January 2021 and is much lower, especially in the second half of 2021. The October and December releases appear quite similar, but if you squint, you can see small differences.

We can calculate the differences:

```{r}
metro_comp |> 
  pivot_wider(names_from = "report_month", values_from = "value") |> 
  mutate(diff_mar_oct = mar_22 - oct_22,
         diff_mar_dec = mar_22 - dec_22,
         diff_oct_dec = oct_22- dec_22) |> 
  pivot_longer(cols = starts_with("diff_"), names_to = "report_date", values_to = "difference") |> 
  filter(report_date == "diff_oct_dec" & date > ymd("2020-01-01")) |> 
  ggplot(aes(date, difference)) +
  geom_col() 
```

Yes, there are differences for 2022, but the differences are fairly small. 

# Conclusion

Adjusted monthly NTD data is indeed adjusted from one month to another. Whether that adjustment is because of data revisions from the reporting agency or something that FTA does (or both) is unclear. The adjustments affect primarily the current year, and they can be large or small. It is possible that the COVID shock to ridership had an outsize influence on the adjustments being made.

Personally, I will be more cautious with current-year data and either not report on it or add least add more cautionary language about data quality. I welcome feedback from others who know more about the NTD reporting and adjustment process.