---
title: "Impervious areas in Madison"
description: |
  Data viz with waffle charts, or not
author:
  - name: Harald Kliems 
    orcid_id: 0000-0003-2542-0047
    url: https://haraldkliems.netlify.app/
date: 2026-04-28
creative_commons: CC BY-SA
categories:
  - Madison (WI)
  - chart
output:
  distill::distill_article:
    self_contained: false
    code_folding: true
---

Madison is having a wet year. The cumulative precipitation for 2026 has been close to [record highs](https://github.com/vgXhc/madison-weather/raw/main/graphs/AnnualCumulativePrecipitation_USW00014837.png), and several rain events led to localized flooding. Impervious surfaces like building roofs or parking lots contribute to this flooding. The city of Madison has an [open data set of those areas](https://data-cityofmadison.opendata.arcgis.com/datasets/2ed52d3fe98740a787c67c9c6af3d0a2_7/about), inspiring me to a little data visualization exercise. 

```{r}
library(tidyverse)

library(waffle)
library(gt)
imp <- read_csv("data/Stormwater_Modeling_Impervious_Land_Cover_Type.csv")
glimpse(imp)
```

First we need to do some cleanup. The `source_area_dest` variable is the human readable description of the type of impervious area. It has `r length(imp$source_area_desc)` values:

```{r}
imp |> 
  distinct(source_area_desc)
```

We consolidate the two different roof types into one and don't distinguish between paved and unpaved parking. The area is provided in square feet and we'll convert to square miles. 

```{r}

df <- imp |> 
  mutate(source_area_desc = fct_recode(
    source_area_desc, 
    Buildings = "Sloped Roof", 
    Buildings = "Flat Roof",
    Parking = "Unpaved Parking"
  )) |> 
  group_by(source_area_desc) |> 
  summarize(area = sum(ShapeSTArea)) |> 
  mutate(area_sq_mi = area * 3.587006e-8)

df
```

This would make for a perfectly fine table (and we'll make that later), but why not have a little fun and turn it into a waffle chart? The [`waffle`](https://github.com/hrbrmstr/waffle/) package makes that relatively easy. "Relatively" because in practice you need to do a lot of tinkering to get a truly nice looking chart. Pick a sensible number of categories, units, comprehensible (but included in the free Fontawesome icon set!) icons, arrange rows and columns, scale everything just right, etc. Here's my solution:

```{r layout="l-body-outset", fig.width = 8, fig.height= 8}
df |> 
  mutate(source_area_desc = if_else(
    source_area_desc %in% (df %>% slice_min(area, n = 7) %>% pull(source_area_desc)),
    "other",
    source_area_desc
  )) %>%
  summarise(area = sum(area_sq_mi*5), .by = source_area_desc) |> 
  arrange(desc(area)) |> 
  mutate(source_area_desc = fct_reorder(source_area_desc, -area)) |> 
  ggplot(aes(label = source_area_desc, values = area)) +
  geom_pictogram(
    size = 8, 
    n_rows = 10,
    aes(colour = source_area_desc),
    flip = TRUE
     #make_proportional = TRUE
  ) +
  scale_label_pictogram(
  name = NULL,
  values = c("building", "road", "parking", "warehouse", "water", "walking", "map"),
  labels = c("Buildings", "Streets", "Parking", "Driveways", "Water body areas", "Sidewalks", "other")
) +
  scale_color_manual(
    name = NULL,
    values = viridisLite::viridis(7, option = "D"),
    labels = c("Buildings", "Streets", "Parking", "Driveways", "Water body areas", "Sidewalks", "other")
  ) +
  coord_equal() +
  hrbrthemes::theme_ipsum_rc(grid="") +
  waffle::theme_enhance_waffle() +
  labs(title = "Impervious surface in Madison",
       subtitle = "Each symbol represents ~1/5 square mile",
       caption = "Data: City of Madison Stormwater Modeling Impervious Land Cover Type") +
  theme(
    legend.key.height = unit(2.25, "line"),
    legend.text = element_text(size = 10, hjust = 0, vjust = 0.75)
  ) 
```

Now as I said, the waffle chart/pictogram looks cool. But if we're being honest, this waffle chart could have been a table. So let's make a nice one. 

```{r}
df |> 
  mutate(source_area_desc = if_else(
    source_area_desc %in% (df %>% slice_min(area, n = 7) %>% pull(source_area_desc)),
    "other",
    source_area_desc
  )) %>%
  mutate(total_area = sum(area)) |> 
  summarise(area_pct = sum(area/total_area), area = sum(area_sq_mi), .by = source_area_desc) |> 
  arrange(desc(area)) |> 
  gt() |> 
  cols_label(source_area_desc = "Type",
             area = html("Total area (mi<sup>2</sup>)"),
             area_pct = "% of total impervious area") |> 
  cols_move(columns = area_pct, after = area) |> 
  fmt_percent(columns = "area_pct",
              decimals = 1) |> 
  fmt_number(columns = area, decimals = 2) |> 
  tab_header(title = paste0("Madison has ", round(df |> summarize(total = sum(area_sq_mi)) |> pull(total), 1), " square miles of impervious surface"),
             subtitle = "Buildings and streets make up over half of that") |> 
  tab_source_note("Data: City of Madison")
  
```

Depending on the audience and the story you're trying to tell, either form of presenting the data could be appropriate. For example, if the exact proportions in each category matter, a table probably is the better choice. But to get a quick glimpse of the major types of impervious surfaces, the waffle chart is a good looking visualization. 
